Constructing Quotient Inductive-Inductive Types

### 2018-10-10 paper.tex
fixes

\[
\Delta^\D\,(N,z,s) \equiv (N^D:N\ra\Set)\times N^D\,z\times ((x:N)\ra N^D\,x\ra N^D\,(s\,x))
\]
The operation $\blank^\D$ is the unary logical predicate
interpretation \cite{bernardy12parametricity}: a context $\Gamma$ is
interpreted as a predicate $\Gamma^\D:\Gamma^\A\ra\Set$. A type
$A:\Ty\,\Gamma$ becomes a predicate depending on a witness of
$\Gamma^\D$, that is, $A^\D:\Gamma^\D\,\gamma\ra
A^\A\,\gamma\ra\Set$. A term $t:\Tm\,\Gamma\,A$ is interpreted as
$t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)$.

@@

The above can be summarized by saying that there is a natural isomorphism
between $\Tms\,\Gamma\,\Delta$ and $\Tm\,\Gamma\,(\K\,\Delta)$. An
alternative definition is given by \emph{democratic} CwFs
\cite{clairambault2014biequivalence}, and it was shown by
\cite{forsberg-phd} that the two definitions are interderivable.

@@

Previously given operations on the theory of signatures yield a
fragment of a full CwF$^\K_{\Eq}$ model; we summarize this below:

\vspace{1em}
\begin{tabular}{l l}
  $\blank^\A$ & \hspace{1em}$\Con$ \\
  $\blank^\D$ & \hspace{1em}$\Ty$  \\
  $\blank^\M$ & \hspace{1em}$\Tms$ \\
  $\blank^\S$ & \hspace{1em}$\Tm$
\end{tabular}
\vspace{1em}

Algebras become the objects of a CwF$^\K_{\Eq}$, homomorphisms become
the morphisms, and displayed algebras and sections together yield
families. However, there are numerous other components in a
CwF$^\K_{\Eq}$ which we will consider in Section \ref{sec:cwfmodel},
where we construct the model.

@@

\textbf{Context extension} is taking total algebras of displayed
algebras. This is analogous to total categories of displayed
categories \cite{displayedCategories}. More concretely, we construct
the $\Sigma$ type of the carrier set $N$ and the family $N^D$, and
glue together the algebra operators with their displayed counterparts.
\begin{alignat*}{5}
  & \blank\rhd \blank : (\Gamma : \Con) \ra \Ty\,\Gamma \ra \Con \\
  & (N, z, s) \rhd (N^D, z^D, s^D) :\equiv
  (((n : N) \times N^D\,n),\,(z,\,z^D),\,(\lambda (x,\,x^D).\,(s\,x,\,s^D\,x\,x^D)))
\end{alignat*}

@@

First, we define initiality and induction in an arbitrary CwF$^\K_{\Eq}$. Below, by $isContr$ we mean unique inhabitation.
\begin{alignat*}{5}
  & Initial : \Con \ra \Set && \hspace{1.5em}Induction : \Con \ra \Set \\
  & Initial\,\Gamma :\equiv (\Delta : \Con) \ra isContr\,(\Tms\,\Gamma\,\Delta) && \hspace{1.5em}Induction\,\Gamma :\equiv (A : \Ty\,\Gamma)\ra\Tm\,\Gamma\,A
\end{alignat*}

@@

In this section we follow the row-based approach and describe the rows
of the model informally, while giving the full definition of the first
4 rows ($\Con$, $\Ty$, $\Tms$, $\Tm$) and the rows for $\U$, $\El$ and
$\Pi$ in Appendix \ref{sec:motives}. These are the most interesting
parts of the construction. The rest of the construction is tedious but
not very enlightening. The first 4 columns are fully formalised in
Agda and most parts of the category-columns from the CwF-columns are
also formalised.

The \textbf{CwF part} of the theory of signatures has to be
interpreted as the CwF of CwF$^\K_{\Eq}$s. Hence, $\Con$ is
interpreted as just CwF$^\K_{\Eq}$, while $\Sub$ is interpreted as the
set of strict CwF$^\K_{\Eq}$-morphisms (which strictly preserve all
structure), $\Ty$ is interpreted as the set of displayed
CwF$^\K_{\Eq}$s, and $\blank\rhd\blank$ is interpreted as the total
CwF$^\K_{\Eq}$ of a displayed one. Previously we illustrated how the
CwF of natural number algebras work in section
\ref{sec:cwfalg}. Constructing the CwF of CwF$^\K_{\Eq}$s is an
analogous, albeit much larger work. All in all, this part is a tedious
although mostly mechanical exercise.

@@

The interpretation of $\Pi\,a\,B:\Ty\,\Gamma$:
{\footnotesize
\begin{alignat*}{5}
  & {(\Pi\,a\,B)}^\A\,\gamma && :\equiv (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) \\
  & {(\Pi\,a\,B)}^\D\,\gamma^D\,f  && :\equiv (\alpha^D:a^\D\,\gamma^D)\ra B^\D\,(\gamma^D,\alpha^D)\,(f\,\alpha) \\
  & {(\Pi\,a\,B)}^\M\,\gamma^M\,f\,f'  && :\equiv (\alpha:a^\A\,\gamma) \ra B^\M\,(\gamma^M,\refl)\,(f\,\alpha)\,(f'\,(a^\M\,\gamma^M\,\alpha)) \\
  & {(\Pi\,a\,B)}^\S\,\gamma^S\,f\,f^D && :\equiv (\alpha:a^\A\,\gamma) \ra B^\S\,(\gamma^S,\refl)\,(f\,\alpha)\,(f^D\,(a^\S\,\gamma^S\,\alpha)) \\
  & \cdot_{\Pi\,a\,B}\,\_  && :\equiv \cdot_B \\
  & (f\rhd_{\Pi\,a\,B} f^D)\,(\alpha,\alpha^D) && :\equiv (f\,\alpha\rhd_B f^D\,\alpha^D) \\
  & (f^D[f^M]_{\Pi\,a\,B})\,\alpha^D && :\equiv f^D\,\alpha^D[f^M\,\alpha]_B \\
  & (\id_{\Pi\,a\,B}\,f)\,\alpha  && :\equiv \id_B\,(f\,\alpha) \\
  & ({f^M}'\circ_{\Pi\,a\,B} f^M)\,\alpha && :\equiv ({f^M}'\,(a^\M\,\gamma^M\,\alpha)\circ_B f^M\,\alpha) \\
  & \epsilon_{\Pi\,a\,B}\,\_  && :\equiv \epsilon_B \\
  & (f^M,_{\Pi\,a\,B} f^S)\,\alpha  && :\equiv (f^M\,\alpha,_B f^S\,\alpha) \\
  & {\pi_1}_{\Pi\,a\,B}\,f^M\,\alpha  && :\equiv {\pi_1}_B\,(f^M\,\alpha) \\
  & {\pi_2}_{\Pi\,a\,B}\,f^M\,\alpha  && :\equiv {\pi_2}_B\,(f^M\,\alpha) \\
  & (f^S[f^M]_{\Pi\,a\,B})\,\alpha && :\equiv f^\S\,(a^\M\,\gamma^M\,\alpha)[f^M\,\alpha]_B \\
  & {[\id]}_{\Pi\,a\,B}  && :\equiv {[\id]}_B \\
  & {[\circ]}_{\Pi\,a\,B}  && :\equiv {[\circ]}_B \\
  & {\ass}_{\Pi\,a\,B}  && :\equiv \ass_B \\
  & {\idl}_{\Pi\,a\,B}  && :\equiv \idl_B \\
  & {\idr}_{\Pi\,a\,B}  && :\equiv \idr_B \\
  & {\cdot\eta}_{\Pi\,a\,B}  && :\equiv {\cdot\eta}_B \\
  & {\rhd\beta_1}_{\Pi\,a\,B}  && :\equiv {\rhd\beta_1}_B \\
  & {\rhd\beta_2}_{\Pi\,a\,B}  && :\equiv {\rhd\beta_2}_B \\
  & {\rhd\eta}_{\Pi\,a\,B}  && :\equiv {\rhd\eta}_B \\
  & {,\circ}_{\Pi\,a\,B}  && :\equiv {,\circ}_B \\
  & \K_{\Pi\,a\,B}\,f\,\alpha  && :\equiv \K_B\,(f\,\alpha) \\
  & {\K[]}_{\Pi\,a\,B}  && :\equiv {\K[]}_B \\
  & \mk_{\Pi\,a\,B}\,f^M\,\alpha && :\equiv \mk_B\,(f^M\,\alpha)\\
  & \unk_{\Pi\,a\,B}\,f^S\,\alpha  && :\equiv \unk_B\,(f^S\,\alpha) \\
  & {\K\beta}_{\Pi\,a\,B}  && :\equiv {\K\beta}_B \\
  & {\K\eta}_{\Pi\,a\,B}  && :\equiv {\K\eta}_B \\
  & {\mk[]}_{\Pi\,a\,B}  && :\equiv {\mk[]}_B \\
  & \Eq_{\Pi\,a\,B}\,f^{S0}\,f^{S1}\,\alpha^D  && : \Eq_B\,(f^{S0}\,\alpha)\,(f^{S1}\,\alpha) \\
  & {\Eq[]}_{\Pi\,a\,B} && :\equiv {\Eq[]}_B \\
  & \eqreflect_{\Pi\,a\,B}\,f^S && :\equiv \funext\,\lambda\alpha.\eqreflect_B\,(f^S\,\alpha)
\end{alignat*}
}

>>>>

\[
\Delta^\D\,(N,z,s) \equiv (N^D:N\ra\Set)\times N^D\,z\times ((x:N)\ra N^D\,x\ra N^D\,(s\,x))
\]
The operation $\blank^\D$ is the unary logical predicate
interpretation \cite{bernardy12parametricity}: a context $\Gamma$ is
interpreted as a predicate $\Gamma^\D:\Gamma^\A\ra\Set$. A type
$A:\Ty\,\Gamma$ becomes a predicate depending on a witness of
$\Gamma^\D$, that is, $A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set$,
where we implicitly quantify over $\gamma$. A term
$t:\Tm\,\Gamma\,A$ is interpreted as
$t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)$.

@@

The above can be summarized by saying that there is a natural isomorphism
between $\Tms\,\Gamma\,\Delta$ and $\Tm\,\Gamma\,(\K\,\Delta)$. An
alternative definition is given by \emph{democratic} CwFs
\cite{clairambault2014biequivalence}, and it was shown in
\cite{forsberg-phd} that the two definitions are interderivable.

@@

Previously given operations on the theory of signatures yield a
fragment of a full CwF$^\K_{\Eq}$ model. Algebras become the objects
of a CwF$^\K_{\Eq}$, homomorphisms become the morphisms, and displayed
algebras and sections together yield families. However, there are
numerous other components in a CwF$^\K_{\Eq}$ which we will consider
in Section \ref{sec:cwfmodel}, where we construct the model.

@@

\textbf{Context extension} is taking total algebras of displayed
algebras. This is analogous to total categories of displayed
categories \cite{displayedCategories}. More concretely, we construct
the $\Sigma$-type of the carrier set $N$ and the family $N^D$, and
glue together the algebra operators with their displayed counterparts.
\begin{alignat*}{5}
  & \blank\rhd \blank : (\Gamma : \Con) \ra \Ty\,\Gamma \ra \Con \\
  & (N, z, s) \rhd (N^D, z^D, s^D) :\equiv
  (((n : N) \times N^D\,n),\,(z,\,z^D),\,(\lambda (x,\,x^D).\,(s\,x,\,s^D\,x\,x^D)))
\end{alignat*}

@@

First, we define initiality and induction in an arbitrary CwF$^\K_{\Eq}$.
\begin{alignat*}{5}
  & Initial : \Con \ra \Set \\
  & Initial\,\Gamma :\equiv (\Delta : \Con) \ra (\sigma : \Tms\,\Gamma\,\Delta)\times((\delta : \Tms\,\Gamma\,\Delta)\ra\sigma=\delta) \\
  & Induction : \Con \ra \Set \\
  & Induction\,\Gamma :\equiv (A : \Ty\,\Gamma)\ra\Tm\,\Gamma\,A
\end{alignat*}

@@

In this section we follow the row-based approach and describe the rows
of the model informally, while giving the full definition of the first
4 rows ($\Con$, $\Ty$, $\Tms$, $\Tm$) and the rows for $\U$, $\El$ and
$\Pi$ in Appendix \ref{sec:motives}. These are the most interesting
parts of the construction. The rest of the construction is tedious and
not very enlightening. The first 4 columns are fully formalised in
Agda and most parts of the category-columns from the CwF-columns are
also formalised.

The \textbf{CwF part} of the theory of signatures has to be
interpreted as the CwF of CwF$^\K_{\Eq}$s. Hence, $\Con$ is
interpreted as just CwF$^\K_{\Eq}$, while $\Sub$ is interpreted as the
set of strict CwF$^\K_{\Eq}$-morphisms (which strictly preserve all
structure), $\Ty$ is interpreted as the set of displayed
CwF$^\K_{\Eq}$s, and $\blank\rhd\blank$ is interpreted as the total
CwF$^\K_{\Eq}$ of a displayed one. Previously we illustrated how the
CwF of natural number algebras work in section
\ref{sec:cwfalg}. Constructing the CwF of CwF$^\K_{\Eq}$s is an
analogous, albeit much larger work. All in all, this part is a mostly
mechanical exercise.

@@

The interpretation of $\Pi\,a\,B:\Ty\,\Gamma$:
{\footnotesize
\begin{alignat*}{5}
  & {(\Pi\,a\,B)}^\A\,\gamma && :\equiv (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) \\
  & {(\Pi\,a\,B)}^\D\,\gamma^D\,f  && :\equiv (\alpha^D:a^\D\,\gamma^D)\ra B^\D\,(\gamma^D,\alpha^D)\,(f\,\alpha) \\
  & {(\Pi\,a\,B)}^\M\,\gamma^M\,f\,f'  && :\equiv (\alpha:a^\A\,\gamma) \ra B^\M\,(\gamma^M,\refl)\,(f\,\alpha)\,(f'\,(a^\M\,\gamma^M\,\alpha)) \\
  & {(\Pi\,a\,B)}^\S\,\gamma^S\,f\,f^D && :\equiv (\alpha:a^\A\,\gamma) \ra B^\S\,(\gamma^S,\refl)\,(f\,\alpha)\,(f^D\,(a^\S\,\gamma^S\,\alpha)) \\
  & \cdot_{\Pi\,a\,B}\,\_  && :\equiv \cdot_B \\
  & (f\rhd_{\Pi\,a\,B} f^D)\,(\alpha,\alpha^D) && :\equiv (f\,\alpha\rhd_B f^D\,\alpha^D) \\
  & (f^D[f^M]_{\Pi\,a\,B})\,\alpha^D && :\equiv f^D\,\alpha^D[f^M\,\alpha]_B \\
  & (\id_{\Pi\,a\,B}\,f)\,\alpha  && :\equiv \id_B\,(f\,\alpha) \\
  & ({f^M}'\circ_{\Pi\,a\,B} f^M)\,\alpha && :\equiv ({f^M}'\,(a^\M\,\gamma^M\,\alpha)\circ_B f^M\,\alpha) \\
  & \epsilon_{\Pi\,a\,B}\,\_  && :\equiv \epsilon_B \\
  & (f^M,_{\Pi\,a\,B} f^S)\,\alpha  && :\equiv (f^M\,\alpha,_B f^S\,\alpha) \\
  & {\pi_1}_{\Pi\,a\,B}\,f^M\,\alpha  && :\equiv {\pi_1}_B\,(f^M\,\alpha) \\
  & {\pi_2}_{\Pi\,a\,B}\,f^M\,\alpha  && :\equiv {\pi_2}_B\,(f^M\,\alpha) \\
  & (f^S[f^M]_{\Pi\,a\,B})\,\alpha && :\equiv f^\S\,(a^\M\,\gamma^M\,\alpha)[f^M\,\alpha]_B \\
  & {[\id]}_{\Pi\,a\,B}  && :\equiv {[\id]}_B \\
  & {[\circ]}_{\Pi\,a\,B}  && :\equiv {[\circ]}_B \\
  & {\ass}_{\Pi\,a\,B}  && :\equiv \ass_B \\
  & {\idl}_{\Pi\,a\,B}  && :\equiv \idl_B \\
  & {\idr}_{\Pi\,a\,B}  && :\equiv \idr_B \\
  & {\cdot\eta}_{\Pi\,a\,B}  && :\equiv {\cdot\eta}_B \\
  & {\rhd\beta_1}_{\Pi\,a\,B}  && :\equiv {\rhd\beta_1}_B \\
  & {\rhd\beta_2}_{\Pi\,a\,B}  && :\equiv {\rhd\beta_2}_B \\
  & {\rhd\eta}_{\Pi\,a\,B}  && :\equiv {\rhd\eta}_B \\
  & {,\circ}_{\Pi\,a\,B}  && :\equiv {,\circ}_B \\
  & \K_{\Pi\,a\,B}\,f\,\alpha  && :\equiv \K_B\,(f\,\alpha) \\
  & {\K[]}_{\Pi\,a\,B}  && :\equiv {\K[]}_B \\
  & \mk_{\Pi\,a\,B}\,f^M\,\alpha && :\equiv \mk_B\,(f^M\,\alpha)\\
  & \unk_{\Pi\,a\,B}\,f^S\,\alpha  && :\equiv \unk_B\,(f^S\,\alpha) \\
  & {\K\beta}_{\Pi\,a\,B}  && :\equiv {\K\beta}_B \\
  & {\K\eta}_{\Pi\,a\,B}  && :\equiv {\K\eta}_B \\
  & {\mk[]}_{\Pi\,a\,B}  && :\equiv {\mk[]}_B \\
  & \Eq_{\Pi\,a\,B}\,f^{S0}\,f^{S1}\,\alpha^D  && : \Eq_B\,(f^{S0}\,\alpha)\,(f^{S1}\,\alpha) \\
  & {\Eq[]}_{\Pi\,a\,B} && :\equiv {\Eq[]}_B \\
  & \eqreflect_{\Pi\,a\,B}\,f^S && :\equiv \funext\,(\lambda\alpha.\,\eqreflect_B\,(f^S\,\alpha))
\end{alignat*}
}
